---
title: "Case Study 1"
author: "Rithvik"
date: "2023-02-15"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ask

## The stakeholders:

-   **Lily Moreno:** The director of marketing and manager

-   **Cyclistic marketing analytics team**

-   **Cyclistic executive team**

## The question to be answered is:

-   *How do annual **members** and **casual** riders use Cyclistic bikes
    **differently**?*

### These questions below are not asked, but any insight will be helpful in that direction as well

-   Why would casual riders buy Cyclistic annual memberships?

-   How can Cyclistic use digital media to influence casual riders to
    become members?

# Prepare

The data is obtained from
[tripdata](https://divvy-tripdata.s3.amazonaws.com/index.html) provided
for this study. We will use only the data for the year 2022, from 1st
Jan 2022 to 31st Dec 2022

For this, the data in csv files is downloaded to local drive. They match
the regular expression pattern "2022.\*divvy.\*.csv":

```{r}
list.files(path = "..", pattern = "2022.*divvy.*.csv",full.names = TRUE)
```

# Process

To begin processing this data, we will first load the packages we need.
The code below attempts to install the packages if it does not find them
in the library, and loads them later

## Reading the data

The data is too large to push to github, so it has been downloaded to
the top-level directory outside git repository.

```{r}
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
if (!require("lubridate")) install.packages("lubridate")
library(lubridate)
```

```{r message=FALSE, warning=FALSE}
tripdata <-
  list.files(path = "..", pattern = "2022.*divvy.*.csv",full.names = TRUE) %>% 
  map_df(~read_csv(.))

```

## Checking the data

Lets look at the data

```{r}
names(tripdata)
```

These are column names

```{r}
str(tripdata)
```

This provides the formats for the data read. As we can see, some columns
are better off as factors

Columns we should have as factors are: rideable_type and member_casual

```{r}
tripdata$rideable_type <- as.factor(tripdata$rideable_type)
table(tripdata$rideable_type)
tripdata$member_casual <- as.factor(tripdata$member_casual)
table(tripdata$member_casual)
```

OK, there are no spurious entries for these factors. The dates are ok.
Are there NA values?

```{r}
sum(is.na(tripdata$started_at))
sum(is.na(tripdata$ended_at))
sum(is.na(tripdata$ride_id))
sum(is.na(tripdata$rideable_type))
sum(is.na(tripdata$start_lat))
sum(is.na(tripdata$end_lat))
sum(is.na(tripdata$start_lng))
sum(is.na(tripdata$end_lng))
```

Only the end_lat and end_lng have an identical number of NAs. Why?

```{r}
table(tripdata[is.na(tripdata$end_lng),]$rideable_type)
```

This really doesn't say much.

## Creating new variables

We need to derive some additional data from the columns already present

1.  Duration of the ride - **duration**

2.  Date, extract from timestamp 'started_at' We will ignore corner
    cases where rides were taken around midnight, and use only
    started_at for the same - **date**

3.  Weekday - Sunday, Monday etc. that the ride was taken -
    **day_of_the_week** - a factor

4.  Whether the ride was a weekday ride(Mon..Fri) or a weekend
    ride(Sat,Sun) - **weekday_or_end** - a factor

5.  A combination of start_station_id and end_station_id - **commute**

For duration, it is simple. Luckily, the started_at and ended_at are
both in POSIX_ct format as timestamps.

```{r}
tripdata$duration <- tripdata$ended_at - tripdata$started_at
```

```{r}
sum(tripdata$duration<0)
sum(tripdata$duration==0)
```

Remove spurious data. Only duration\>0 should be included. The 100
negative entries are likely errors in entry.

```{r}
tripdata[tripdata$duration==0,]
```

Its not clear why there are entries for 0 sec durations as they are
either maintenance calls or errors. So we will remove all these:

```{r}
tripdata <- tripdata[tripdata$duration>0,]
```

Date can easily be extracted, and the

```{r}
tripdata$date <- lubridate::date(tripdata$started_at)
```

From this, we get the day_of_the_week

```{r}
tripdata$day_of_the_week <- as.factor(wday(tripdata$started_at, label=TRUE))
```

We now need to decide whether the day is a weekday or weekend. This is
simple:

```{r}
weekday_or_end <- function(day) ifelse(day=="Sat"|day=="Sun","Weekend","Weekday") 
```

We define the simple one-line function. Then we plug it into the
assignment below. Note that these values should be factors

```{r}
tripdata$weekday_or_end <- as.factor(weekday_or_end(as.character(tripdata$day_of_the_week)))
```

One last check

```{r}
str(tripdata)
```

OK

# Analyse

## Extracting information

For analysis, we can generate a large number of graphs, only some of
which may make sense, while others may give the same basic information.
The various factor variables are member_casual, weekday_or_end,
rideable_type and day_of_the_week. All of these can be used for
group_by. The member_casual factor is key, as the question to be
answered revolves around it. So we will generate group-based analysis
for each of the other factors. The value we need to check is duration,
as it is the billable quantity here. We take sum, mean, mode, max, min
and count.

### What day of the week

```{r}
trips_by_weekday <- tripdata %>%
                group_by(day_of_the_week, member_casual,date) %>%
                summarise(sum = sum(duration), mean = mean(duration), mode = mode(duration), max = max(duration), min = min(duration), count = sum(duration>10,na.rm=TRUE))
```

### Weekly trips on working day or weekend?

```{r}
weekly_trips <- tripdata %>%
                group_by(weekday_or_end, member_casual,date) %>%
                summarise(sum = sum(duration), mean = mean(duration), mode = mode(duration), max = max(duration), min = min(duration), count = sum(duration>10,na.rm=TRUE))
```

### What type of bike is preferred?

```{r}
bike_prefs <- tripdata %>%
                group_by(rideable_type, member_casual,date) %>%
                summarise(sum = sum(duration), mean = mean(duration), mode = mode(duration), max = max(duration), min = min(duration),count = sum(duration>10,na.rm=TRUE))
```

Many plots can be generated. But which of them make sense, which of them
give us the information we need?

### Data for weekdays

#### Count

```{r}

gg_weekday_bar <- ggplot(data = trips_by_weekday, aes(x=day_of_the_week, y=count, fill=member_casual)) +
            facet_grid(.~member_casual) +
            geom_col()+
            ggtitle("Week day durations - Count")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekday_bar)
```

We can see that there are more member rides than casual rides. The
casual riders are more in the weekends. We will see more weekday vs
weekend information later. \#### Sum

```{r}

gg_weekday_bar <- ggplot(data = trips_by_weekday, aes(x=day_of_the_week, y=sum, fill=member_casual)) +
            facet_grid(.~member_casual) +
            geom_col()+
            ggtitle("Week day durations - Sum")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekday_bar)
```

The pattern is even stronger here: in weekends, casual riders go out for
long durations or many trips

#### Mean

```{r}

gg_weekday_bar <- ggplot(data = trips_by_weekday, aes(x=day_of_the_week, y=mean, fill=member_casual)) +
            facet_grid(.~member_casual) +
            geom_col()+
  ggtitle("Week day durations - Mean")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekday_bar)
```

Looks like casual riders ride for longer durations. Lets confirm this:

#### Max

```{r}

gg_weekday_bar <- ggplot(data = trips_by_weekday, aes(x=day_of_the_week, y=max, fill=member_casual)) +
            facet_grid(.~member_casual) +
            geom_col()+
            ggtitle("Week day durations - Max")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekday_bar)
```

Looks like casual riders love to go on long rides. Not necessarily in
the weekend, but certainly more. Member riders however seem to use these
rides for fixed durations. How consistent in this are they?

### Weekdays vs Weekends

We will now plot timelines for all the data through the year and get
trends.

#### Sum

```{r}
gg_weekly_sums <- ggplot(data = weekly_trips, aes(x=date, y=sum, col=member_casual)) +
            facet_grid(weekday_or_end~member_casual) +
            geom_point() +
            geom_smooth()+
  ggtitle("Timeline trend - Duration sum")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekly_sums)
```

The members are much more consistent. Obviously the usage is more in the
summer months.

#### Mean

```{r}
gg_weekly_means <- ggplot(data = weekly_trips, aes(x=date, y=mean, col=member_casual)) +
            facet_grid(weekday_or_end~member_casual) +
            geom_point() +
            geom_smooth()+
ggtitle("Timeline trend - Duration mean")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekly_means)
```

This strongly reinforces the consistency of the members

#### Count

```{r}
g <- ggplot(data = weekly_trips, aes(x=date, y=count, col=member_casual)) +
            facet_grid(weekday_or_end~member_casual) +
            geom_point() +
            geom_smooth()+
ggtitle("Timeline trend - Duration count")+
            theme(plot.title = element_text(hjust = 0.5))
print(g)
```

There is not much appreciable difference in counts. It seems members and
casual riders take similar number of rides in the weekdays and weekends.

#### Max

```{r}
gg_week_max <- ggplot(data = weekly_trips, aes(x=date, y=max, col=member_casual)) +
            facet_grid(weekday_or_end~member_casual) +
            geom_point() +
            geom_smooth()+
ggtitle("Timeline trend - Duration max")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_week_max)
```

This shows a drastic difference. The duration of member rides is highly
restricted and those of casual riders are not. One obvious reason is
that members use the rides for daily commute to work, while casual
riders use them for adventure or recreation. To confirm this theory,
lets take a histogram of means

```{r}
gg_weekday_hist <- ggplot(data = weekly_trips, aes(x=mean, fill=member_casual)) +
            facet_grid(weekday_or_end~member_casual) +
            geom_histogram()+
    ggtitle("Histogram of mean duration")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_weekday_hist)
```

This tells us that short-distance commutes are what members prefer,
while casual riders frequently go on long duration trips.

### Type of Bike

```{r}
gg_bike_counts <- ggplot(data = bike_prefs, aes(x=date, y=count, col=member_casual)) +
            facet_grid(rideable_type~member_casual) +
            geom_point() +
            geom_smooth() +
  ggtitle("Timeline trend - Bike type sum")+
            theme(plot.title = element_text(hjust = 0.5))
print(gg_bike_counts)
```

There is no obvious difference in usage of type of bike, though there is
a preference by casual riders towards classic bike. There is no data for
docked bike for members, docked bikes are used exclusively by casual
users.

```{r}

gg_bike_sum <- ggplot(data = bike_prefs, aes(x=date, y=sum, col=member_casual)) +
            facet_grid(rideable_type~member_casual) +
            geom_point() +
            geom_smooth()
print(gg_bike_sum)
```

### Commute data

Lets check the type of rides that are preferred by casual vs members
First lets remove NAs on start station and end station ids:

```{r}
tripdata_filtered <- tripdata[!is.na(tripdata$start_station_id) & !is.na(tripdata$end_station_id),]
```

Now, we make a combined variable of start station and end station. If
start station and end station are identical, it means a round trip. We
can have a boolean variable for that.

```{r}
tripdata_filtered$commute_from_to <- as.factor(paste(tripdata_filtered$start_station_name,tripdata_filtered$end_station_name,sep=".to."))
tripdata_filtered$commuteid_from_to <- as.factor(paste(tripdata_filtered$start_station_id,tripdata_filtered$end_station_id,sep="_-_"))
tripdata_filtered$roundtrip <- tripdata_filtered$start_station_id == tripdata_filtered$end_station_id

```

We can now get sum , mean, max and count elements for

```{r}
tripdata_filtered$commuteid_from_to <- as.factor(tripdata_filtered$commuteid_from_to)

commute_prefs <- tripdata_filtered %>%
                  group_by(member_casual,roundtrip) %>%
                  summarise(sum = sum(duration), mean=mean(duration),max=max(duration),count = sum(duration>10,na.rm=TRUE)) 
commute_stats <- tripdata_filtered %>%
                  group_by(commuteid_from_to,member_casual,roundtrip) %>%
                  summarise(count = sum(duration>10,na.rm=TRUE))
## since there are a large number of commuteid_from_to groups, don't use sum in summarise, or it takes forever
```

```{r}
commute_stats_top <- commute_stats %>%
                      arrange(desc(count)) %>%
                      group_by(member_casual) %>%
                      slice(1:10)
commute_stats_top
```

For casual riders, most of the rides are round trips, where destination
is same as starting point.

```{r}
gg_commute_bar <- ggplot(data = commute_prefs, aes( x=member_casual, y=mean, fill=roundtrip)) +
#            facet_grid(.~roundtrip) +
            geom_col()+
            ggtitle("Commute Preferences - Roundtrip (TRUE/FALSE)")+
            theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
print(gg_commute_bar)
```

This shows casual users prefer to use round trips for fairly long
commutes.

# Conclusions

Using the above charts we can come to the conclusions that:-

1.  Casual users use bikes more frequently in the weekends than in the
    weekdays

2.  Casual users prefer long duration trips, while members have fixed
    duration trips

3.  Member trip durations are very consistent - they indicate possible
    commute times to and from work. Casual users use rides likely for
    recreational activities or visits.

4.  Casual members use more round trips. However, members also utilize
    round trips, though lesser. This may indicate errand runs during
    work.
